---
title: "Class05: Data Viz with GGPLOT"
author: "Kavi (PID: A69046927)"
format: gfm
---

Today we are playing and plotting graphics with R.
 
There are lots of ways to make cool figures in R. There is "base" R graphics (`plot()`, `hist()`, `boxplot()`, etc.)

There is also add-on packages, like **ggplot**

```{r}
head(cars)
```
Let's plot this with "base" R:

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg,mtcars$disp, pch=2)
```
```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 package is **ggplot()**. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`. 

> **N.B.** I never want to run `install.packages()` in my quarto source document!!!


```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed,y=dist) +
  geom_point()
```
```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with `ggplot(cars)`)
- The **aesthetics** mapping (given with `aes()`)
- The **geom** (given with `geom_point()`)

> For simple canned graphs, "base" R is nearly always faster

### Adding more layers

Let's add a line and a title, subtitle and caption as well as custom axis labels

```{r}
ggplot(cars) +
  aes(x=speed,y=dist) +
  geom_point() +
  geom_smooth(method="lm",se=FALSE) +
  labs(title = "Silly Plot",
       x = "Speed (MPH)",
       y = "Distance (ft)",
       caption="when is tea time anyway???") +
  theme_bw()
```

## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset

> Q2. How many "up" regulated genes are there?


```{r}
sum(genes$State == "up")
```
There are `r sum(genes$State == "up")` "up" regulated genes.

```{r}
table(genes$State)
```

```{r}
p<-ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point() +
    scale_color_manual(values=c("purple","orange","black"))
p
```

```{r}
p + labs(title="look at me")
```

Silly example of adding labels
```{r}
library(ggrepel)
pg<-ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State, label=Gene) +
    geom_text(color="gray") +
    geom_point() +
    scale_color_manual(values=c("purple","orange","black")) +
    geom_text_repel(max.overlaps=100) +
    theme_classic()
pg
```

## Going Further

Playing with different dataset to make something animated

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```
```{r}
tail(gapminder)
```

Classic plot of increasing GDP overtime.

```{r}
ggplot(gapminder) +
         aes(x=gdpPercap,y=lifeExp, col=continent, size=gdpPercap) +
         geom_point(alpha=0.2) +
        facet_wrap(~continent) +
        theme_bw()
```
Wanted to try spitting more bars... couldn't
```{r}
ggplot(gapminder) +
   aes(x=lifeExp) +
  geom_bar()
```


## Bonus gganimate code that never rendered lol

library(gapminder)
library(gganimate)

# Setup nice regular ggplot of the gapminder data
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Facet by continent
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

