---
title: "Find a Gene Project"
author: "Kavi (PID: )"
toc: true
format: html
---

> 7. Convert the alignment to FASTA if needed and use R/Bio3D to calculate a sequence identity matrix.Generate a sequence identity heatmap in R and include it in the report.

```{r, echo=TRUE}
library(bio3d)

aln <- read.fasta("find-a-gene.fa")
idmat <- seqidentity(aln)

# widen plot margins so labels aren't cut off
par(mar = c(12, 12, 4, 4))  # bottom, left, top, right

heatmap(idmat,
        symm = TRUE,
        Rowv = NA,
        Colv = NA,
        scale = "none"
,
        cexRow = 0.9,
        cexCol = 0.9,
        margins = c(12,12),
        las = 2,         # rotate labels for readability
        main = "Sequence Identity Heatmap")

# restore defaults
par(mar = c(5,4,4,2))
```
> 8. Use Bio3D or online BLAST to search the PDB for structural homologs of your sequence. Report the top 3 unique PDB hits with E-value, sequence identity, PDB ID, experimental method, resolution, and source organism.

# Generating consensus sequence
```{r}
library(bio3d)
cons_aln <- consensus(aln)
```

# Performing BLAST search against PDB
```{r}
aln_blast <- blast.pdb(cons_aln$seq)
plot.blast(aln_blast,
       top = 10,
       main = "BLAST Search Results against PDB")
```
# Extracting top 3 unique PDB hits
```{r}
hits <- aln_blast$hit.tbl
hits <- hits[order(hits$evalue), ]
hits$pdb4 <- substr(hits$pdb.id, 1, 4)
all_pdb4 <- unique(hits$pdb4)
all_ann <- pdb.annotate(all_pdb4,
                            anno.terms = c("structureId",
                                            "experimentalTechnique",
                                            "resolution",
                                            "source"))
all_ann_source <- all_ann[!duplicated(all_ann$source), ]
hits_sub <- hits[hits$pdb4 %in% all_ann_source$structureId, ]
all_ann_source_hits <- merge(all_ann_source, hits_sub,
                             by.x = "structureId",
                             by.y = "pdb4")
all_ann_source_hits <- all_ann_source_hits[order(all_ann_source_hits$evalue), ]
```

# Annotating the top 3 structures
```{r}
top3 <- head(all_ann_source_hits, 3)
top3
```

# Combined info

```{r}
result <- data.frame(
  pdb.id     = top3$pdb.id,
  evalue     = top3$evalue,
  identity   = top3$identity,
  method     = top3$experimentalTechnique,
  resolution = top3$resolution,
  source     = top3$source
)

result
```
# Write a CSV of the result
```{r}
write.csv(result, "top3_pdb_hits.csv", row.names = FALSE)
```

# Looking for conserved residues
```{r}
library(bio3d)
cons <- conserv(aln$ali)
plot(cons)
which(cons > 0.8)
```
```{r}
# aln$ali is alignment matrix
query_index <- 3   # whichever column corresponds to your D. melanogaster Su(var)3-3
query_seq <- aln$ali[, query_index]

high <- which(cons >= 0.8)

# keep only positions where your query sequence is NOT a gap
high_query_positions <- high[ query_seq[high] != "-" ]

high_query_positions[1:500]
```
```{r}
alignment_positions <- high_query_positions

# alignment column → residue number
res_nums <- cumsum(query_seq != "-")[alignment_positions]

res_nums[1:10]
```
```{r}
aln$id
sapply(1:length(aln$id), function(i) sum(aln$ali[,i] != "-"))
apply(aln$ali[1:20, ], 2, paste0, collapse="")
```

# Last resort from ChatGPT
```{r}
library(bio3d)

# 1. Read alignment
aln <- read.fasta("find-a-gene.fa")

# 2. Replace any NA with "-" globally
aln$ali[is.na(aln$ali)] <- "-"

# 3. Pick your D. psuedoobscura sequence column
mel_idx <- grep("psuedoobscura", aln$id, ignore.case = TRUE)[1]
query_seq <- aln$ali[, mel_idx]

# sanity check
table(query_seq)   # should show letters + "-" only

```

