---
title: "Class 16 HW"
author: "Kavi (PID: A69046927)"
format: gfm
toc: true
---
Back on our laptop we can now use R and Bioconductor tools to further explore this large scale dataset.

For example there is an R function called tximport() in the tximport package, which enables straightforward import of Kallisto results

With each sample having its own directory containing the Kallisto output, we can import the transcript count estimates into R using:
```{r}
library(tximport)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
```{r}
head(txi.kallisto$counts)
```
We now have our estimated transcript counts for each sample in R. We can see how many transcripts we have for each sample, and how many transcripts are detected in at least one sample:
```{r}
colSums(txi.kallisto$counts)
sum(rowSums(txi.kallisto$counts)>0)
```
Before subsequent analysis, we might want to filter out those annotated transcripts with no reads:
```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```
And those with no change over the samples:
```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```
## Principal Component Analysis
We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:
```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```
Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:
```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
plot(pca$x[,2], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC2", ylab="PC3", pch=16)
plot(pca$x[,1], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC3", pch=16)
```
```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()

ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()

ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```
## Differential Expression Analysis
We can use DESeq2 to complete the differential-expression analysis that we are already familiar with:

An example of creating a DESeqDataSet for use with DESeq2:
```{r,message=F}
library(DESeq2)
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
dds <- DESeq(dds)
res <- results(dds)
head(res)
```
```{r}
my_cols <- rep("gray", nrow(res))
my_cols[which(res$padj < 0.05 & res$log2FoldChange > 1)] <- "red"
my_cols[which(res$padj < 0.05 & res$log2FoldChange < -1)] <- "blue"
```

```{r}
library(ggplot2)
library(ggrepel)

ggplot(res, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(col = my_cols, alpha = 0.6) +
  theme_classic() +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") + 
  labs(x = "Log2 Fold Change", y = "-Log10 P-value")
  # Only label significant hits:
  geom_text_repel(
    data = subset(res, padj < 0.05 & abs(log2FoldChange) > 1),
    aes(label = genename),
    size = 3
  )
```



