---
title: "Class 19"
author: "Kavi Gonur (PID: A69046927)"
toc: true
format: html
---

## Background

Pertussis (a.k.a Whooping Cough) is a highly infectious lung infection caused by the bacteia *B. pertussis*.

The CDC tracks case numbers in the US and makes this data available online:

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, include=F}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063)
       )
```

```{r}
library(ggplot2)

cdcgraph <- ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  labs(x="Year",y="Number of cases") +
  theme_classic()

cdcgraph
```
> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
cdcgraph +
  geom_vline(xintercept = 1946, linetype="dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype="dashed", color = "red") +
  geom_vline(xintercept = 2020, linetype="dashed", color = "grey") +
  geom_text(aes(x=1946, y=250000, label="wP"), 
            angle=90, vjust = -0.5, color="blue") +
  geom_text(aes(x=1996, y=250000, label="aP"), 
            angle=90, vjust = -0.5, color="red")
```
> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Maybe the vaccine wasn't as effective as hoped, or there were changes in vaccination rates, pathogen evolution, or reporting practices. Further investigation would be needed to determine the exact cause.

## The CMI-PB Project

The CMI-PB project is a collaboration between researchers at UCSD and the Scripps Institution of Oceanography to study the microbial communities in the coastal waters of Southern California. The project involves collecting water samples from various locations along the coast and analyzing the microbial DNA using high-throughput sequencing techniques.

They make their data aavailable via a JSON format running API. E=We can read JSON format with the `read_json` function from the jsonlite R package..
```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```
> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
subject$infancy_vac
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
Let's read more tables 
```{r}
library(jsonlite)
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",simplifyVector = TRUE)
```

Join (or link, or merge) using the 

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```
```{r}
ab_data <- inner_join(meta,ab_titer)
head(ab_data)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
head(ab_data$isotype)
```
> How many different antigens are there in the dataset?

```{r}
unique(ab_data$antigen)

```
```{r}
ggplot(ab_data) +
  aes(MFI,antigen) +
  geom_boxplot() +
  theme_classic()
```
## Focus in IgG

IgG is crucial for long-term immunity and responding to bacterial and viral infections

```{r}
ab_data |>
  filter(isotype == "IgG") -> igg_data
igg_data
```
Plot of antigen levels again but for IgG only
```{r}
igg_dataplot <- ggplot(igg_data) +
  aes(x=MFI_normalised, y=antigen) +
  geom_boxplot() +
  theme_classic()
igg_dataplot
```

### Differences between aP and wP?

We can color up by the `infancy_vac` values of "wP" or "aP"

```{r}
igg_dataplot +
  aes(color=infancy_vac)
```
We could "facet" by the "aP" vs "wP" column

```{r}
igg_dataplot +
  aes(color=infancy_vac) +
  facet_wrap(~infancy_vac)
```

### Time course analysis

We can use `visit` as a proxy for time here and facet our plots by this value 1 to 8...

```{r}
igg_data |>
  filter(visit %in% 1:8) |>
  ggplot() +
    aes(x=MFI_normalised,y=antigen,color=infancy_vac) +
    facet_wrap(~visit) +
    geom_boxplot() +
    theme_classic()
```

## Time course of PT (Virulence Factor: Pertussis Toxin)
```{r}
pt <- igg_data |>
  filter(antigen == "PT") |>
  filter(dataset == "2021_dataset")

pt
```

```{r}
pt |>
  ggplot() +
    aes(planned_day_relative_to_boost,
        MFI_normalised, 
        color=infancy_vac, 
        group=subject_id) +
    geom_point() +
    geom_line() +
    theme_classic() +
    geom_vline(xintercept =14, col="darkgreen") +
    labs(x="Days post booster", y="Normalized MFI for PT IgG")
```
## System setup
```{r}
sessionInfo()
```

