---
title: "Class 19"
author: "Kavi Gonur (PID: A69046927)"
toc: true
format: html
---

## Background

Pertussis (a.k.a Whooping Cough) is a highly infectious lung infection caused by the bacteia *B. pertussis*.

The CDC tracks case numbers in the US and makes this data available online:

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, include=F}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063)
       )
```

```{r}
library(ggplot2)

cdcgraph <- ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  labs(x="Year",y="Number of cases") +
  theme_classic()

cdcgraph
```
> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
cdcgraph +
  geom_vline(xintercept = 1946, linetype="dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype="dashed", color = "red") +
  geom_vline(xintercept = 2020, linetype="dashed", color = "grey") +
  geom_text(aes(x=1946, y=250000, label="wP"), 
            angle=90, vjust = -0.5, color="blue") +
  geom_text(aes(x=1996, y=250000, label="aP"), 
            angle=90, vjust = -0.5, color="red")
```
> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Maybe the vaccine wasn't as effective as hoped, or there were changes in vaccination rates, pathogen evolution, or reporting practices. Further investigation would be needed to determine the exact cause.

## The CMI-PB Project

The CMI-PB project is a collaboration between researchers at UCSD and the Scripps Institution of Oceanography to study the microbial communities in the coastal waters of Southern California. The project involves collecting water samples from various locations along the coast and analyzing the microbial DNA using high-throughput sequencing techniques.

They make their data aavailable via a JSON format running API. E=We can read JSON format with the `read_json` function from the jsonlite R package..
```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```
> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
subject$infancy_vac
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
Let's read more tables 
```{r}
library(jsonlite)
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",simplifyVector = TRUE)
```

### Working with Dates
> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
library(dplyr)

subject$age <- today() - ymd(subject$year_of_birth)

# (i)
ap <- subject %>% filter(infancy_vac == "aP")
round(summary(time_length(ap$age, "years" )))

 # (ii)
wp <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length(wp$age, "years")))

# (iii)
t.test(ap$age,wp$age)
```
i) 28
ii) 37
iii) yes, significantly different (p 2.2e-16)

> Q8. Determine the age of all individuals at time of boost?

```{r}
subject$boost_age <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
round(head(time_length(subject$boost_age,"years")))
round(summary(time_length(subject$boost_age,"years")))
```

> Q9a. With the help of a faceted boxplot or histogram, do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```
I think they are significantly different!

Join (or link, or merge) using the 

> Q9b. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:
> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```
```{r}
ab_data <- inner_join(meta,ab_titer)
head(ab_data)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
head(ab_data$isotype)
```
> How many different antigens are there in the dataset?

```{r}
unique(ab_data$antigen)
```
```{r}
ggplot(ab_data) +
  aes(MFI,antigen) +
  geom_boxplot() +
  theme_classic()
```
> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(ab_data$dataset)
```
There's a lot more rows in the most recent dataset!

## Focus in IgG

IgG is crucial for long-term immunity and responding to bacterial and viral infections

```{r}
ab_data |>
  filter(isotype == "IgG") -> igg_data
```
Plot of antigen levels again but for IgG only
```{r}
igg_dataplot <- ggplot(igg_data) +
  aes(x=MFI_normalised, y=antigen) +
  geom_boxplot() +
  theme_classic()
igg_dataplot
```

### Differences between aP and wP?

We can color up by the `infancy_vac` values of "wP" or "aP"

```{r}
igg_dataplot +
  aes(color=infancy_vac)
```
We could "facet" by the "aP" vs "wP" column

```{r}
igg_dataplot +
  aes(color=infancy_vac) +
  facet_wrap(~infancy_vac)
```

### Time course analysis

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

We can use `visit` as a proxy for time here and facet our plots by this value 1 to 8...

```{r}
igg_data |>
  filter(visit %in% 1:8) |>
  ggplot() +
    aes(x=MFI_normalised,y=antigen,color=infancy_vac) +
    facet_wrap(~visit) +
    geom_boxplot() +
    theme_classic()
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
library(dplyr)
filter(igg_data, antigen=="PRN") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_classic() +
  labs(title="PRN")

filter(igg_data, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_classic() +
  labs(title="FIM2/3")
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

Of the data presented in the example: PT levels overtime rise and exceed OVA. 
In this dataset, FIM levels start out large but drop significantly more than PRN (this could be becuse FIM just had a greater range of values than PRN.)

> Q17. Do you see any clear difference in aP vs. wP responses?

PRN: Responses about the same
FIM: For most part, aP > wP.

## Time course of PT (Virulence Factor: Pertussis Toxin)
(removed dataset displays here, made rendering take WAYYYY too long)
```{r}
pt_2020 <- igg_data |>
  filter(antigen == "PT") |>
  filter(dataset == "2020_dataset")
pt_2021 <- igg_data |>
  filter(antigen == "PT") |>
  filter(dataset == "2021_dataset")
```

```{r}
pt_2020 |>
  ggplot() +
    aes(planned_day_relative_to_boost,
        MFI_normalised, 
        color=infancy_vac, 
        group=subject_id) +
    geom_point() +
    geom_line() +
    theme_classic() +
    geom_vline(xintercept=0, col="darkgreen") +
    geom_vline(xintercept =14, col="darkgreen") +
    labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)",x="Days post booster", y="Normalized MFI for PT IgG")

pt_2020 |>
  ggplot() +
    aes(planned_day_relative_to_boost,
        MFI_normalised, 
        color=infancy_vac, 
        group=subject_id) +
    geom_point() +
    geom_line() +
    theme_classic() +
    geom_vline(xintercept=0, col="darkgreen") +
    geom_vline(xintercept =14, col="darkgreen") +
    scale_x_continuous(limits = c(0, 125)) +
    labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels), \n scale cropped at 125 days for better comparison with 2021",x="Days post booster", y="Normalized MFI for PT IgG")

pt_2021 |>
  ggplot() +
    aes(planned_day_relative_to_boost,
        MFI_normalised, 
        color=infancy_vac, 
        group=subject_id) +
    geom_point() +
    geom_line() +
    theme_classic() +
    geom_vline(xintercept=0, col="darkgreen") +
    geom_vline(xintercept =14, col="darkgreen") +
    labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)",x="Days post booster", y="Normalized MFI for PT IgG")
```
> Q18. Does this trend look similar for the 2020 dataset?

Hard to tell - the 2020 trend looks quite smushed bnecause of twice the number of days being recorded. I would guess probably.

Actually, no, MFIs were HIGHER in 2020 at 14 days.

## System setup
```{r}
sessionInfo()
```