---
title: "Class 10"
author: "Kavi Gonur (PID: A69046927)"
format: gfm
---

## Importing candy data

```{r}
candy_file <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv")

candy = data.frame(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` different candy types in the dataset.

> Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` fruity candy types in the dataset.

## What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

I have a ton of favorite candies. How dare you make me choose. ;-; But I suppose right now I like fruitier candy so Haribo Happy Cola. Its winpercent is `r candy["Haribo Happy Cola", ]$winpercent`%.

> Q4. What is the winpercent value for “Kit Kat”?

`r candy["Kit Kat", ]$winpercent`%

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

`r candy["Tootsie Roll Snack Bars", ]$winpercent`%

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes. winpercent.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

0 = candy is NOT a chocolate. 1 = candy IS a chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) +
  aes(x=winpercent) +
  theme_classic() +
  geom_histogram(bins=20) +
  labs(x="Win Percent", y= "Count")
```
> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(x=winpercent) +
  theme_classic() +
  geom_density()
```
Nope

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
According to the mean, above. But the median is below.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
choc.candy
# 2. Extract their `winpercent` values
choc.win <- choc.candy$winpercent

# 3. Find the mean of these values
choc.mean <- mean(choc.win)

# 4-6. Do the same for fruity candy
fruity.inds <- as.logical(candy$fruity)
fruity.candy <- candy[fruity.inds,]
fruity.win <- fruity.candy$winpercent
fruity.mean <- mean(fruity.win)
fruity.mean
# 7. Which mean value is higher?
if (fruity.mean > choc.mean) {
  print("Fruity")
} else {
  print("Chocolate")
}
```


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,fruity.win)
```

## Overall Candy Rankings
> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% 
arrange(winpercent) %>% 
head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)
candy %>% 
arrange(winpercent) %>% 
tail(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```
### Time to add some useful color
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent
```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c("pricepercent", "winpercent")], n=5 )
```
Tootsie Roll Midgies

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c("pricepercent", "winpercent")], n=5 )
```
Nik L Nips

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
## Exploring the correlation structure
```{r}
library(corrplot)
cij <- cor(candy)
cij
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

- Fruity + chocolate/caramel/peanutyalmondy/nougat/crispedricewafer/bar

> Q23. Similarly, what two variables are most positively correlated?

- chocolate + bar or chocolate + winpercent

## Principal Component Analysis
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
ggplot(pca$x) +
  aes(PC1,PC2) +
  geom_point(col=my_cols) +
  theme_classic() +
  geom_text_repel(aes(label = rownames(pca$x)), color = my_cols, size = 3.3, max.overlaps = 5)
```
```{r}
ggplot(pca$rotation) +
  aes(PC1,rownames(pca$rotation))
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


