---
title: "Class 17"
author: "Kavi (PID: A69046927)"
format: html
editor: visual
---
## ENSEMBL/OMIM Calculations
```{r}
library(dplyr)
read.csv("MXL_genotypes.csv") -> MXL_genotypes
table(MXL_genotypes)
sum(MXL_genotypes$Genotype..forward.strand.=="G|G") -> homoz_numb
all_numb = count(MXL_genotypes)
homoz_prop <- (homoz_numb/all_numb) * 100
homoz_prop
```
## DESeq Analysis
```{r}
genereads <- read.table("genereads.txt", header=TRUE, row.names=1)

summary(genereads)

aa <- sum(genereads$geno == "A/A")
ag <- sum(genereads$geno == "A/G")
gg <- sum(genereads$geno == "G/G")
```
> Q13 Sample Size

There are ```r aa``` individuals with A/A genotype, ```r ag``` individuals with A/G genotype, and ```r gg``` individuals with G/G genotype.

Median expression provided in boxplot for Q14.

> Q14 Box Plot

```{r}
library(ggplot2)
ggplot(genereads) +
  aes(x=geno,y=exp) +
  theme_classic() +
  geom_boxplot() +
   stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(..y.., 2)),
    vjust = -0.5,
    color = "red"
  ) +
  labs(x="Genotype", y="Expression Level") +
  geom_jitter(alpha=0.2, width = 0.2)
```
> Response to Q14

From the boxplot, I can infer that the more copies there are of the SNP allele variant, the lower the expression of ORMDL3. The allele inhibits proper expression of ORMDL3.
